---
title: 'Class 1: An introduction to Missing Data Analysis'
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{maynooth_uni_logo.jpg}
  \newline \vspace{1cm}
  https://andrewcparnell.github.io/mda_course 
  \newline PRESS RECORD 
output:
  beamer_presentation:
    includes:
      in_header: header.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf')
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
```

## Let's get started

- Introduction from Novartis?

- Timetable for the week

- Pre-requisites

## How this course works

- This course lives on GitHub, which means anyone can see the slides, code, etc, and make comments on it
- The timetable document (`index.html`) provides links to all the pdf slides and practicals
- The slides and the practicals are all written in `Rmarkdown` format, which means you can load them up in Rstudio and see how everything was created
- Let me know if you spot mistakes, as these can be easily updated on the GitHub page
- There is a `mda_course.Rproj` R project file from which you should be able to run all the code

## Copyright statement

All the non-GitHub materials provided in the Introduction to Missing Data Analysis course are copyright of Andrew Parnell.

This means:

- As a user (the student) you have permission (licence) to access the materials to aid and
support your individual studies.
- You are not permitted to copy or distribute any materials without the relevant permission
- As faculty we may reserve the right to remove a user in the event of any possible infringement

## Course format and other details

- We do two lectures in the morning (separated by a coffee break) then have a practical class where we go through some R code together
- There is also an (optional) exercise sheet with further questions for those who are particularly keen
- If you want to send me a private message use the message board and I will try to answer them as we go
- Please ask lots of questions, but __MUTE YOUR MICROPHONE__ when not asking them
- Some good books:

    - _Statistical Analysis with Missing Data_ by Little and Rubin 
    - _Bayesian Data Analysis_ by Gelman et al (chapter on Missing Data Analysis)
    - 


## Why care about missing data analysis?



## Structure of this class


## A simple example

```{r, echo = FALSE, fig.height = 5}
# Feed conversion rate example
plot(1,1)
```

## Missing data analysis terrible jargon

MCAR
MAR
NMAR (MNAR)

## Other jargon

imputation
amputation(!)
monotone missing
ignorable
non-ignorable
Complete case (Ch 3)
Available case (p61/73)

## Missing data analysis common notation


## Other examples where missing data occurs


## Bivariate MAR

book P22

## How people usually analyse missing data

Book P23 (p35)

## Available case vs complete case analysis

Complete case (Ch 3)
Available case (p61/73)

## Simple imputation

P67 Ch 4

## Explicit vs implicit imputation

P68/80

## A warning about imputation

P67 Ch 4

## A simple demonstration of the pifalls of imputation

P81/69

## Regression imputation

P71/83

## General guidelines on imputation

P81/93

## Estimating uncertainties with imputed data

P90/102

## From simple to multiple imputation

P95/107



